---
title: "Shiny App - US states spatial analysis"
author: "Alejandro Barrero Bilbao"
date: '`r Sys.Date()`'
output:
  html_document: 
    toc: true
    toc_float: true
    keep_md: true
runtime: shiny
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```
### Introduction

This R Markdown report concisely summarises the functionalities of a basic app based on the `spData::us_states` dataset. 

For further information about this library and its datasets for spatial analysis, please refer to [the official library site](https://nowosad.github.io/spData/).

The `us_states` dataset comprises the following features for all 48 *continental* states of the US and the District of Columbia, whereby the term *continental* is loosely defined here as referring to all states in mainland America between Canada and Mexico, hence excluding Hawaii and Alaska:

* `GEOID`: character vector of geographic identifiers
* `NAME`: character vector of state names
* `REGION`: character vector of region names
* `AREA`: area in square kilometers of units class
* `total_pop_10`: numerical vector of total population in 2010
* `total_pop_15`: numerical vector of total population in 2015
* `geometry`: sfc_MULTIPOLYGON

Check that the required packages are installed/loaded successfully:
```{r loading, message=FALSE, echo=c(7,11)}

# Remove all existing objects from the current environment
rm(list = ls())

# Install (if necessary) and load required packages
install_packages_if_not_present <- function(x) {
    if (sum(!x %in% installed.packages())) {
        install.packages(x[!x %in% installed.packages()])
    }
}
packages <- c("spData",
              "leaflet",
              "shiny",
              "data.table")
install_packages_if_not_present(packages)

# Show loaded packages
sapply(packages, require, character.only = TRUE)
```
### Aggregation by state/region

The most basic subset of the reference dataset is by state `NAME` and `REGION`. In the interactive maps below, you can select your state of interest (tab `NAME`) and its relevant features for display in the `data.table` header. Alternatively, click on the highlighted state to have its summarised features pop up.


Additionally, you can select your region of interest (tab `REGION`) for it to be highlighted in the map and its cumulative and average features per state summarised in the `data.table` header. Hovering over the states in the highlighted region enables the display of their name.

```{r app1}

shinyApp(
  
  ui <- fluidPage(
          tags$head(tags$style('.selectize-dropdown {z-index: 10000}')),
      
          tabsetPanel(
            
              tabPanel("NAME",
                        fluidRow(
                            column(6,
                                   selectInput("states", label = h4("Select a state"),
                                                choices = us_states$NAME,
                                                selected = us_states$NAME[29],
                                                width='100%',
                                                multiple = F)
                            ),
                            column(6,
                                   selectInput("stfeat", label = h4("Select state feature(s)"),
                                                choices = colnames(us_states)[2:6],
                                                selected = colnames(us_states)[2],
                                                width='100%',
                                                multiple = T)
                            )
                        ),
                        DT::dataTableOutput("state_dt"),
                        leafletOutput("myUSmap_byState")
              ),
          
              tabPanel("REGION",
                        selectInput("region", label = h4("Select a region"),
                                     choices = us_states$REGION,
                                     selected = 1,
                                     width='50%',
                                     multiple = F),
                        DT::dataTableOutput("region_dt"),
                        leafletOutput("myUSmap_byRegion")
              )
          )
  ),
  
  server <- function(input, output) {
    data.index <- reactive({
        which(us_states$NAME == input$states)
    })
    
    region.states <- reactive({
        subset(us_states, REGION == input$region, NAME, drop = T)
    })
    
    data.longitude <- reactive({
        us_states$geometry[[data.index()]][[1]][[1]][,1]
    })
    
    data.latitude <- reactive({
        us_states$geometry[[data.index()]][[1]][[1]][,2]
    })
    
    data.popup <- reactive({
        paste('State: ', input$states, '<br>',
              'Population 2010: ', us_states$total_pop_10[data.index()], '<br>',
              'Population 2015: ', us_states$total_pop_15[data.index()], '<br>',
              'Region: ', us_states$REGION[data.index()], '<br>',
              'Area: ', us_states$AREA[data.index()])
    })
    
    
    output$myUSmap_byState <- renderLeaflet({
        
        mymap <- leaflet(options = leafletOptions(minZoom = 3, maxZoom = 5)) %>%
            setView(lng = -98.5795, lat = 37.8283, zoom = 4) %>%
            addProviderTiles(providers$CartoDB.Positron) %>%
            addPolygons(lng = data.longitude(), lat = data.latitude(), weight = 1.5, popup = data.popup())
        
        # Accommodate multi-polygons
        if (length(us_states$geometry[[data.index()]]) > 1) {
            
            for (pol in 1:length(us_states$geometry[[data.index()]])) {
                lng <- us_states$geometry[[data.index()]][[pol]][[1]][,1]
                lat <- us_states$geometry[[data.index()]][[pol]][[1]][,2]
                
                mymap <- addPolygons(mymap, lng = lng, lat = lat, weight = 1.5, popup = data.popup())
            }
        }
        return(mymap)
    })
    
    compute_st_dt <- reactive({
        dat <- us_states[data.index(), input$stfeat]
        dat <- as.data.table(dat)
        return(dat)
    })
    
    output$state_dt <- DT::renderDataTable(compute_st_dt(), rownames=FALSE)
    
    compute_rg_dt <- reactive({
        dat <- subset(us_states, REGION == input$region, c("REGION", "AREA", "total_pop_15"))
        dat <- as.data.table(dat)
        dat <- dat[, .(Tot.Area = sum(AREA),
                       Avg.Area = mean(AREA),
                       Tot.Population15 = sum(total_pop_15),
                       Avg.Population15 = mean(total_pop_15)), by=REGION]
        return(dat)
    })
    
    output$region_dt <- DT::renderDataTable(compute_rg_dt(), rownames=FALSE)
    
    output$myUSmap_byRegion <- renderLeaflet({
        
        mymap <- leaflet(options = leafletOptions(minZoom = 3, maxZoom = 5)) %>%
            setView(lng = -98.5795, lat = 37.8283, zoom = 4) %>%
            addProviderTiles(providers$CartoDB.Positron)
            
        
        for (s in region.states()) {
          
            index <- which(us_states$NAME == s)
            lng <- us_states$geometry[[index]][[1]][[1]][,1]
            lat <- us_states$geometry[[index]][[1]][[1]][,2]
            
            mymap <- addPolygons(mymap, lng = lng, lat = lat,
                                 weight = 1.5, label = us_states$NAME[index])
            
            # Accommodate multi-polygons
            if (length(us_states$geometry[[index]]) > 1) {
            
                for (pol in 1:length(us_states$geometry[[index]])) {
                  
                    lng <- us_states$geometry[[index]][[pol]][[1]][,1]
                    lat <- us_states$geometry[[index]][[pol]][[1]][,2]
                    
                    mymap <- addPolygons(mymap, lng = lng, lat = lat,
                                         weight = 1.5, label = us_states$NAME[index])
                }
            }
        }
        return(mymap)
    })
  },
  
  options = list(height = 700, width = '100%')
)
```


### Evolution of population 2010-2015

The only demographic features available in the `us_states` dataset are the total population values in years 2010 and 2015. With these, you can inspect the relative change in population over the timespan or, combined with each state `AREA`, the population density for either year.


You can focus on a specific `REGION` or on the entire country from the dropdown menus below. Hovering over the active `REGION` enables the display of its state `NAME`s, and clicking on any state activates a popup with the demographic metric of interest.

```{r app2}

shinyApp(
  
  ui <- fluidPage(
            tags$head(tags$style('.selectize-dropdown {z-index: 10000}')),
            fluidRow(
                            column(6,
                                   selectInput("region", label = h4("Select a region"),
                                                choices = c("All", levels(us_states$REGION)),
                                                selected = 1,
                                                width='100%',
                                                multiple = F)
                            ),
                            column(6,
                                   selectInput("rgdmft", label = h4("Select region demographic feature"),
                                                choices = c("Population change 2010-2015",
                                                            "Population density 2010",
                                                            "Population density 2015"),
                                                selected = 1,
                                                width='100%',
                                                multiple = F)
                            )
                        ),
            leafletOutput("population_byRegion")
        ),
  
  server <- function (input, output) {
    
      region.states <- reactive({
          if (input$region == "All") {
              rg <- subset(us_states, TRUE, NAME, drop = T)
          } else {
              rg <- subset(us_states, REGION == input$region, NAME, drop = T)
          }
          return(rg)
      })
      
      region.pop <- reactive({
          subset(us_states, NAME %in% region.states(), c(total_pop_15, total_pop_10))
      })
      
      region.geo <- reactive({
          subset(us_states, NAME %in% region.states(), geometry, drop = T)
      })
      
      pop.change <- reactive({
          (region.pop()$total_pop_15 - region.pop()$total_pop_10)/region.pop()$total_pop_10 * 100
      })

      pop.ft <- reactive({
          if (input$rgdmft == "Population change 2010-2015") {
              pf <- (us_states$total_pop_15 - us_states$total_pop_10)/us_states$total_pop_10 * 100
          } else if (input$rgdmft == "Population density 2010") {
              pf <- us_states$total_pop_10/as.numeric(us_states$AREA)
          } else {
              pf <- us_states$total_pop_15/as.numeric(us_states$AREA)
          }
          return(pf)
      })
      
      color.palette <- reactive({
          if (input$rgdmft == "Population change 2010-2015") {
              pal <- colorNumeric(palette = "Blues", domain = pop.ft())
              cp <- pal(pop.ft())
          } else {
              # Hard code the District of Columbia to be encoded as black
              # (it is an outlier and messes up the colour scale)
              i <- which.max(pop.ft())
              pal <- colorNumeric(palette = "Blues", domain = pop.ft()[-i])
              cp <- append(pal(pop.ft()[-i]), "#000000", i-1)
          }
          return(cp)
      })
      
      palette.domain <- reactive({
          if (input$rgdmft == "Population change 2010-2015") {
              pd <- pop.ft()
          } else {
              # Exclude the District of Columbia from the legend colour gradient
              # (it is an outlier and messes up the colour scale)
              i <- which.max(pop.ft())
              pd <- pop.ft()[-i]
          }
          return(pd)
      })
      
      popup.long <- reactive({
          if (input$rgdmft == "Population change 2010-2015") {
              pul <- 'Population change [%]: '
          } else if (input$rgdmft == "Population density 2010") {
              pul <- 'Population density 2010 [pp/km2]: '
          } else {
              pul <- 'Population density 2015 [pp/km2]: '
          }
          return(pul)
      })
      
      popup.short <- reactive({
          if (input$rgdmft == "Population change 2010-2015") {
              pus <- 'Pop.change [%]'
          } else if (input$rgdmft == "Population density 2010") {
              pus <- 'Pop.density 2010'
          } else {
              pus <- 'Pop.density 2015'
          }
          return(pus)
      })
    
      output$population_byRegion <- renderLeaflet({
        
          mymap <- leaflet(options = leafletOptions(minZoom = 3, maxZoom = 5)) %>%
                   setView(lng = -100.5795, lat = 37.8283, zoom = 4) %>%
                   addProviderTiles(providers$CartoDB.Positron)
        
          for (s in region.states()) {
            
              index <- which(region.states() == s)
              abs.index <- which(us_states$NAME == s)
              lng <- region.geo()[[index]][[1]][[1]][,1]
              lat <- region.geo()[[index]][[1]][[1]][,2]
              
              mymap <- addPolygons(mymap, lng = lng, lat = lat,
                                   weight = 1.5, label = region.states()[index],
                                   color = color.palette()[abs.index],
                                   fillOpacity = 0.5,
                                   popup = paste0(popup.long(),
                                                  round(pop.ft()[abs.index], 2)))
              
              # Accommodate multi-polygons
              if (length(region.geo()[[index]]) > 1) {
              
                  for (pol in 1:length(region.geo()[[index]])) {
                    
                      lng <- region.geo()[[index]][[pol]][[1]][,1]
                      lat <- region.geo()[[index]][[pol]][[1]][,2]
                      
                      mymap <- addPolygons(mymap, lng = lng, lat = lat,
                                           weight = 1.5, label = region.states()[index],
                                           color = color.palette()[abs.index],
                                           fillOpacity = 0.5,
                                           popup = paste0(popup.long(),
                                                          round(pop.ft()[abs.index], 2)))
                  }
              }
          }
          mymap <- addLegend(mymap, "bottomleft",
                             pal = colorNumeric(palette = "Blues", domain = palette.domain()),
                             values = palette.domain(),
                             title = popup.short(),
                             opacity = 0.5)
          
          return(mymap)
      })
  },
  
  options = list(height = 540, width = '100%')
)
```

The **District of Columbia** is an outlier in terms of population density, which is probably related to it being the location of the nation's capital, hence concentrating a large population in about `r round(us_states$AREA[29], 0)` km^2^. Indeed, the capital's density is about 9 times larger than the second most dense state (**Rhode Island**). For this reason, the **District of Columbia** is not considered for the colour palette of the polygons infill colour or the legend colour gradient.


The density maps confirm a historical salient feature of the country's demographics: the north-eastern coast concentrates significantly more population than the other regions. States like **California** and **Illinois** have moderate densities despite embedding the second (*Los Angeles*) and third (*Chicago*) largest cities, respectively.


For the vast majority of states, the trend over the timespan from 2010 to 2015 has been to increase its population, the only two exceptions being **Michigan** and **Rhode Island** (whereby the latter still retains a high density nonetheless). Aside from the **District of Columbia**, the states with the greatest increase in population are large states from regions other than the north-east, such as **North Dakota**, **Utah** or **Texas**, all with relative increase values of about `r ceiling(((us_states$total_pop_15 - us_states$total_pop_10)/us_states$total_pop_10 * 100)[23])`%.